\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-26 doc/incertae_sedis.tex
Move some things around.

In order to determine how the exclude sets should be modified to account for
\emph{incertae sedis} taxa, we first attempt to express the unmodified exclude
sets in a helpful way.  We note that a node $n$ ordinarily excludes the
descendants of $\mathcal{I}(s)$ for all of its siblings, as well as the siblings
of all of its ancestors.  This leads to the following equation:
\begin{align*}
\mathcal{E}(n) & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a)\right]\\
 & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a)\right]\\
 & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\mathcal{E}(parent(n)) \label{eq:exclude-set-formula-traditional}
\end{align*}

When modifying the exclude set for a node $n$, we seek to find all the \emph{incertae
sedis} taxa that may intrude into it. It turns out that the incertae sedis taxa that
might intrude are just the \emph{incertae sedis} siblings of $n$ and all of its ancestors.
\begin{align*}
\mathcal{E}(n) & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\\
 & =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\\
 & \,\,\,\,\,\,\,\,\,\,\,\,\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\\
 & =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\mathcal{E}(parent(n)).\label{eq:exclude-set-formula-1a}
\end{align*}
This formula allows us to compute exclude sets via a pre-order traversal on the
taxonomy tree with exclude set for the root node defined to be $\emptyset$.

@@

@@

\subsection{Cases that this semantics can handle}

\subsection{Surprising results of the split-based semantics}

@@

\subsection{Placement causes broken taxa}
\label{conflict}
Synthesis with \emph{incertae sedis} taxa has the potential to resolve
uncertain taxon placements using information from phylogenies.
Since
phylogenies are ranked higher than taxonomy, they may place members of
incertae sedis clades as they wish even when we do not take into
account the semantics of incertae sedis taxa.
However, without the
incertae sedis semantics described above, placing a taxon $A$ within a
taxon $B$ results in conflict with taxon $B$ in the taxonomy.
This has
three main effects: \begin{enumerate}
    \item The edge leading to clade $B$ is lost.
    \item Any taxonomy-only children of $B$ are placed at the
        MRCA of the phylogenetically informed taxa, instead of at node $B$.
    \item The name for taxon $B$ is lost.
    \item We do not record $A$ as being placed within $B$.
\end{enumerate} In order to avoid a situation
where input phylogenies conflict with a large number of taxa when
\emph{incertae sedis} taxa are placed within them, we have previously
filtered \emph{incertae sedis} taxa from the taxonomy when
constructing all prior synthesis trees.

When the synthesis tree conflicts with a taxonomy node, we say that
the taxon $B$ at that node is a broken taxon.
Broken taxa have two
main effects, both of which are negative.
First, the name $B$ of the
broken taxon is removed from the synthesis tree.
Second, the
conflicting edge for taxon $B$ is not included in the synthesis tree.
This means that any children of $B$ that are taxonomy-only will not be
placed with the children of $B$ that are mentioned in input
phylogenies.
Instead the taxonomy-only children of a broken taxon move
towards the root of the synthesis tree and attach at the first higher-
ranked taxon that is an ancestor of $B$ but is not broken.

We therefore seek a synthesis method that can correctly introduce
\emph{incertae sedis} taxa into containing taxa without breaking the
containing taxa.
We change the semantics of names to imply, not the
exclusion of all non-included taxa, but only some non-included taxa,
as described in section \ref{sec:Semantics-of-incertae}.
As a result
of this change in semantics, placing an IS taxon $A$ within a sister
taxon $B$ no longer results in conflict with $B$.
This allows us to
retain the split for $B$ within the synthesis tree, so that taxonomy-
only children of $B$ are correctly grouped with their siblings that
are referenced by the input trees.
We may then retain the name for the
no-longer-broken taxon.
Finally, we are then able to stop filtering
\emph{incertae sedis} taxa, so that they appear in the synthesis tree.
Thus, the synthesis tree is able to represent substantially more
species, without suffering the loss of taxa and the loss of structure.

@@

\subsection{Case 3: Placement of one \emph{incertae sedis} taxon within
another.}

>>>>

In order to determine how the exclude sets should be modified to account for
\emph{incertae sedis} taxa, we first attempt to express the unmodified exclude
sets in a helpful way.  We note that a node $n$ ordinarily excludes the
descendants of $\mathcal{I}(s)$ for all of its siblings, as well as the siblings
of all of its ancestors.  This leads to the following equation:
\begin{align*}
  \mathcal{E}(n) & = \mathcal{L} - \mathcal{I}(n)\\
 & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a)\right]\\
 & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a)\right]\\
 & =\left[\mathcal{I}(s)\big|s\in siblings(n)\right]\cup\mathcal{E}(parent(n)) \label{eq:exclude-set-formula-traditional}
\end{align*}

When modifying the exclude set for a node $n$, we seek to find all the \emph{incertae
sedis} taxa that may intrude into it. It turns out that the incertae sedis taxa that
might intrude are just the \emph{incertae sedis} siblings of $n$ and all of its ancestors.
\begin{align*}
\mathcal{E}(n) & =\left[\mathcal{I}(s)\big|a\in n+ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\\
 & =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\\
 & \,\,\,\,\,\,\,\,\,\,\,\,\left[\mathcal{I}(s)\big|a\in ancestors(n),s\in siblings(a),s\text{ not \emph{incertae sedis}}\right]\\
 & =\left[\mathcal{I}(s)\big|s\in siblings(n),s\text{ not \emph{incertae sedis}}\right]\cup\mathcal{E}(parent(n)).\label{eq:exclude-set-formula-1a}
\end{align*}
This formula allows us to compute exclude sets via a pre-order traversal on the
taxonomy tree with the exclude set for the root node defined to be $\emptyset$.

@@

\bdrcomment{Ben: discuss naming + interdigitation}

@@

@@

@@

\subsection{Case 3: Placement of one \emph{incertae sedis} taxon within
another.}
\subsection{Surprising results of the split-based semantics}
